---
title: "Base `R` plotting"
author: "Chris Hamm"
date: "`r format(Sys.Date())`"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", cache = TRUE)
```

## Introduction

Base `R` refers to the default suite of tools that come with a standard `R` installation. There are many packages that allow the user to plot in different ways. For example, the package `ggplot2` is particularly popular. However, these alternative packages require learning a new language (the **gg** in `ggplot2` stands for **g**rammer of **g**raphics) and you may not feel inclined to learn another language. Whether you prefer `ggplot2` or base `R` is besides the point. If you are making reproducible figures in `R` you are already winning. For the record, I use both base `R` and `ggplot2`. Both have their own strengths and weaknesses.

We will be working with the`iris` data set, which comes with `R` when you install the software. 

### `iris` data set 

The history of the [`iris` flower dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set) is very interesting. In short, the data were collected by E. Anderson and introduced by one of the founders of statistics, R.A. Fisher, in the [paper](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-1809.1936.tb02137.x/abstract;jsessionid=872BB90F822BC6CED8F4696E544EF6F5.f03t04) that described [`linear discriminant analysis`](https://en.wikipedia.org/wiki/Linear_discriminant_analysis). But enough about that, let's load the data and generate some plots. 

To load the `iris` data set:
```{r load_iris}
data(iris)
```

Nothing appears to have happened but in fact we did load the data set into `R`'s memory. Let's look at the dimensions of the data set and the first 6 rows of data:

```{r}
dim(iris)
head(iris)
```

We can see there are 150 rows and 5 columns of data. Those 5 columns are: 

1. `Sepal.Length`
1. `Sepal.Width`
1. `Petal.Length`
1. `Petal.Width`
1. `Species`

It looks like the first four columns of data should be continuous numbers and the last column should be a factor. Let's make sure that `R` is treating the data accordingly:

```{r check}
str(iris)
```

We are good to go. Let's make some plots.

## Plots

### Histogram

The basic command to create a histogram is `hist()`. We need to specify which of the columns we want to plot in our histogram. Let's plot a histogram of `Sepal.Length`.

```{r hist_1}
hist(iris$Sepal.Length)
```

Yay!

Wait. This is kind of boring to me. 

The real power of plotting in `R` comes from the ability to customize graphics. We plotted with the default parameters for the `hist()` function.

To see the default parameters of a function, and all of the other arguments that can be passed through a function, we use the `?` in front of the function name.

```{r hist_help}
?hist
```

This will open the `Help` window on your computer and show you the arguments and defaults for a function. The `Help` page for a function is the first place I go to troubleshoot an issue.

Let's make the histogram look a little bit nicer. The first thing I want to do is color in the bars of the histogram so thet stand out more. The `col` argument takes a color inside of `""`.

```{r hist_1}

```